Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Give a full explanation of the admissibility of i-pialt early on

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

\Rref{I-Pi} is the most interesting. To explain what it means, we
need to perform a few steps of transformations to its precondition.
The precondition consists of a
mysterious relation $[[R]]$ over the set $[[S]]$, an interpretation
of the type $[[A]]$, and returns a subset of terms. $[[R]]$ is further
subject to the following two constraints:
\begin{itemize}
\item $[[forall a, (# exists S0  , ( a , S0 ) in R #)]]$
\item $[[forall a, forall S0, (# ( a , S0 ) in R implies Interp I i B { a } S0 #)]]$
\end{itemize}
The first condition allows us to induce a function $[[F in S ->
PowerSet STm]]$ such that $[[(a , F a) in R]]$. Then we can easily
show that $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$.

For the other direction, given a function $[[F]]$ satisfying the
property $[[forall a, (# a in S implies Interp I i B { a } F a #) ]]$, it
is possible to  define a relation $[[R]]=[[
{ ( a , F a )  | a in S }]]$. It is easy to verify that $[[R]]$ satisfy
the the constraint required in the precondition of \rref{I-Pi}.

Furthermore, the skolemization process allows us to conclude the
following equivalence:
\[ \exists F, [[forall a, (# a in S implies Interp I i B { a } F a #)]] \iff [[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]  \]
% Ignoring the content below the horizontal bar of
% the derivation, we claim the statements from the precondition are equivalent to
% the conjunction of the following statements:
% \begin{itemize}
% \item $[[Interp I i A S]]$
% \item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
%   [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
% \end{itemize}
% In general, let $R$ be a binary relation over the sets $[[A]]$ and
% $[[B]]$, it easy to verify the following equivalence, which is used to
% justify the skolemization process~\citep{skolemization}.
% \scw{Is this true constructively? If not, is that ok?}
% \[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
%   \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
%   \in R\]
% The left-hand side is equivalent to the right-hand side, but has no
% mentioning of any functions.
The precondition of \rref{I-Pi} therefore can be formulated as the
following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $[[forall a, (# a in S implies (# exists S0 , Interp I i B {a} S0 #) #)]]$
\end{itemize}
From this alternative formulation, we can more easily tell what is
happening in \rref{I-Pi}: the
function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, forall S0, (# ( a , S0 ) in R implies b a in S0 #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[R in S * PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. With the relation $[[R]]$ in scope, we can refer to the
witness sets $[[S0]]$ such that $[[Interp I i B {a} S0]]$ for $[[a in S]]$.
With the alternative representation that is free of the relation $[[R]]$, we
might want to reformulate our logical relation in the following more
concise but equivalent form, whose admissibility follows directly from
\rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since it not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when written as an endofunction over the
domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt}.

@@

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

@@

  All cases are trivial except for the \rref{I-Set} case, where we
  want to show that the set $[[I j]]$ is closed under expansion for
  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
  $[[A in I j]]$ if only if there exists some $[[S]]$ such that
  $[[Interp I i A S]]$. By \rref{I-Par}, we must also have $[[B in I
  j]]$ for all $[[B => A]]$.
\end{proof}

@@

% \begin{lemma}[Adequacy: $CR_1$ for type]
%   \label{lemma:cr1ty}
%   If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
% \end{lemma}
% \begin{proof}
%   Immediately from \rref{I-Ne} and \rref{I-Par}.
% \end{proof}

@@

In terms of our proof technique, we close over our semantically valid types
and terms in \rref{I-Par, I-Bool} using the non-deterministic parallel
reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. Our use of a non-deterministic reduction strategy
means that we need confluence to prove the functionality of our logical
relation. However, the benefit is that it immediately gives us a semantic
justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.

>>>>

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation of
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the relation $[[R in S * PowerSet STm]]$ to be the relation
$[[{ (a , S0 ) | a in S , Interp I  i B { a } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the existence of a
fixed point with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the inversion
principle (Lemma~\ref{lemma:piinvalt}) we derive after proving functionality.

@@

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Red} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

@@

  All cases are trivial except for the \rref{I-Set} case, where we
  want to show that the set $[[I j]]$ is closed under expansion for
  all $[[j < i]]$. However, by the definition of $[[I]]$, we know that
  $[[A in I j]]$ if only if there exists some $[[S]]$ such that
  $[[Interp I i A S]]$. By \rref{I-Red}, we must also have $[[B in I
  j]]$ for all $[[B => A]]$.
\end{proof}

@@

% \begin{lemma}[Adequacy: $CR_1$ for type]
%   \label{lemma:cr1ty}
%   If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
% \end{lemma}
% \begin{proof}
%   Immediately from \rref{I-Ne} and \rref{I-Red}.
% \end{proof}

@@

In terms of our proof technique, we close over our semantically valid types
and terms in \rref{I-Red, I-Bool} using the non-deterministic parallel
reduction relation, while \citet{decagda,nbeincoq} employ a deterministic weak
head reduction relation. Our use of a non-deterministic reduction strategy
means that we need confluence to prove the functionality of our logical
relation. However, the benefit is that it immediately gives us a semantic
justification of the conversion rule (Lemma~\ref{lemma:logrelcoherence}).
% In particular, we obtain the confluence
% result of reduction at a very early stage before we even define our
% logical relation.
